\bf\mftitle

### 2011-01-21 doc/jInfer/twostep/twostep.tex
3159165 Create documentation fof TwoStep simplifier - middlework, added image for factory pattern

\newcommand{\myscale}{0.74}
\newcommand{\vect}[1]{\boldsymbol{#1}}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\jmodule}[1]{\emph{#1}}

@@

\begin{document}
\maketitle
Target audience: developers willing to extend jInfer, scientist willing to implement own inference methods\\

\emph{Note: we use the term \textbf{inference} for the act of creation of schema throughout this and other jInfer documents.}

\section{Using of factory pattern}
TwoStep simplifier is divided into three submodules, which are then divided into submodules and so on.
Each submodule is simply class, properly anotated.
User selects proper classes - submodules in chain to work on inference.
Classes are then, in runtime looked up by using NetBeans lookup mechanism.
But NetBeans holds one instance of each class, that may be looked upon.
When user runs inference first time, classes are looked up and used.
Maybe class members are initialised, instances are in some state after inference completes.
Then, user clicks run button again, and same instances of classes are returned by lookups.
Ooops, theese are not freshly created instances, using them as they are may cause harm.
And problem arises - force each class (submodule) that is being looked up in inference process, to implement
some sort of cleanup method, that would restart it into fresh state and make it ready to use by another 
inference run? Or create new instances of theese classes in each inference run and make singleton classes (those which NetBeans return by lookups) their factory classes?

We decided to use second approach. 
It enables us not only to produce fresh instance in each inference run, but also factory classes implement obligatory
module methods such as \code{getName}, \code{getDescription} and so on, which are really same in each inference run.
Each submodule is then defined by (at least) two interfaces.
One is the factory interface, like this one:
\begin{verbatim}
public interface AutomatonSimplifierFactory extends NamedModule, Capabilities, UserModuleDescription {
  <T> AutomatonSimplifier<T> create();
}
\end{verbatim}
It extends \code{NamedModule} with all module methods.
It extends \code{Capabilities}, thus each module have to answer, if it has some capabilities.
And, in TwoStep we often extend \code{UserModuleDescription}, which defines method for obtaining description of the module, comprehensive to user (it is displayed in properties panels).
In some circumstances, it is useful to have method \code{create} generic.
The automaton simplifier works with automaton, which itself is generic too.
But simplifying does not depend on type of symbol of automaton, so interface \code{AutomatonSimplifier<T>}
is also generic as simplifier can simplify automaton of any symbol java type.
Factory interface deals with this by defining create method generic too.

Second interface is the one, which is returned by \code{create} method in factory.
That is the interface, which defines real work cycle with submodule (in example it is \code{AutomatonSimplifier<T>}).
We call this interface the worker interface of submodule.
Usage of this factory pattern follows the routine:
\begin{verbatim}
final Properties p = RunningProject.getActiveProjectProps(getName());

@@

  return ModuleSelectionHelper.lookupImpl(AutomatonSimplifierFactory.class,
            p.getProperty(PROPERTIES_AUTOMATON_SIMPLIFIER));
}
\end{verbatim}
Cluster processor AutomatonMergingState then receives factories for AutomatonSimplifier and RegexpAutomatonSipmlifier submodules in its constructor.
Cluster processor then may create as many instances of submodule classes as it needs (maybe simplifying more than one automaton).
Thorough this document, we will mention only worker interface when describing submodules, since all factory interfaces are designed same way just described.

@@

    // 3. process rules
    final ClusterProcessor<AbstractStructuralNode> processor= clusterProcessorFactory.create();
    for (Cluster<AbstractStructuralNode> cluster : clusterer.getClusters()) {
      final AbstractStructuralNode node =  processor.processCluster(clusterer, cluster.getMembers() - without sentinels);
      // 4. add to rules
      finalGrammar.add(
              new Element(node.getContext(),
              node.getName(),
              node.getMetadata(),
              ((Element) node).getSubnodes(),
              attList);
    }
\end{verbatim}
Task of clustering is dedicated to \jmodule{Clusterer} submodule, and task of infering regular expression for each cluster is dedicated to \jmodule{ClusterProcessor} submodule. We will examine both of them.
There is also third submodule called \jmodule{RegularExpressionCleaner} actually, its purpose is just to beatify output regular expressions, no inference logic is implemented there. 
Modules currently in jInfer are drawn on fig. \ref{twostep_modules}.
\begin{figure}
	\centering\includegraphics[scale=\myscale]{twostep_modules}
	\caption{Submodules of TwoStep simplifier} \label{twostep_modules}
\end{figure}
We provide one \jmodule{Clusterer} and four \jmodule{ClusterProcessor} implementations.
Each of those will be explained further in this document.

@@

We provide \code{Clusterer} interface for classes to implement.
Its purpose is to cluster bunch of elements (rules) on input, into  bunch of cluster class instances  (cluters).
It has methods \code{add()} and \code{addAll()} for adding items for clustering.
Center is method \code{cluster()}, which does the clustering itself.
As it may be time-consuming operation, method throws \code{InterruptedException}, implementation should take care 
of checking whether thread is user interrupted (see \cite{arch}).
After clustering, implementation should hold clusters in member, as it it will be further asked by calling method \code{T getRepresentantForItem(T item)}.
Given item to this method, one can ask for representant of cluster, to which the item belongs.
If no such cluster exists (item was not added for clustering before), we recommend you throwing an exception rather than returning \code{null}.
Missing will probably  indicate error in algorithm rather than normal workflow.
One can pull clusters from clusterer by calling \code{getClusters()} method.

@@

\subsubsection{ClustererWithAttributes extended interface}
Maybe you noticed that whole clusterer interface and cluster class are generic.
They may be used as design pattern not only for clustering elements in inference process.
To address clustering of elements in more detail, we created \code{ClustererWithAttributes<T, S>} interface, which extends \code{Clusterer<T>} interface.
It add method \code{List<Cluster<S>> getAttributeClusters(T representant)}, implying that each representant of
type \code{T} (that is representant of some main cluster) has some "attribute" clusters associated with it.
Theese clusters are obtained by calling \code{getAttributeClusters(x)}.
Attribute clusters are of type \code{S}.

@@

Whole submodule structure of AutomatonMergingState cluster processor is drawn on fig. \ref{automaton_merging_state_submodules}.
\begin{figure}
	\centering\includegraphics[scale=\myscale]{automaton_merging_state_submodules}
	\caption{Submodules of AutomatonMergingState cluster processor} \label{automaton_merging_state_submodules}
\end{figure}
We have implemented \emph{k,h-context} (see \cite{ahonen}) state equivalence in class \code{KHContext}, which is used by \class{Greedy}
to test mergability of states by default configuration.

\subsubsubsection{StateRemoval}
We are using state removal method (see \cite{1224321}) to convert regexp automaton into equivalent regular expression.
This is implemented in \class{StateRemoval} class.
We defined one submodule of this class with interface called \code{Orderer}.
It has only one method to implement: \code{getStateToRemove}.
Given automaton it has to return reference to one state which should be removed from automaton at first.
State removal calls this submodule and removes states given until there are only two states in automaton - superInitial and superFinal states,
with exactly one transition.
That transition has final regular expression on it as symbol, it is read and returned to AutomatonMergingState.
We implement one orderer, called \code{Weighted}.
It is simple heuristic - weights all states (weight = sum of {in | out | loop}-transition regular expression lengths) and returns state with lowest weight.

>>>>

\newcommand{\myscale}{0.74}
\newcommand{\vect}[1]{\boldsymbol{#1}}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\jmodule}[1]{\texttt{\textsl{#1}}}

@@

\begin{document}
\maketitle
\noindent Target audience: developers willing to extend jInfer, scientist willing to implement own inference methods

\noindent \emph{Note: we use the term \textbf{inference} for the act of creation of schema throughout this and other jInfer documents.}

\section{Using of factory pattern}
\begin{figure}
	\centering\includegraphics[scale=\myscale]{factory_pattern1}
	\caption{Same instance of class returned by NetBeans in two successive inference runs} \label{factory_pattern1}
\end{figure}
\jmodule{TwoStepSimplifier} is divided into three submodules, which are then divided into submodules and so on.
Each submodule is simply class, properly anotated.
User selects proper classes - submodules to work in chain on inference.
Classes are then, in runtime looked up by using NetBeans lookup mechanism (see \cite[p. 16]{arch}).
But NetBeans holds one instance of each class, that may be looked upon.
When user runs inference first time, classes are looked up and used.
Maybe class members are initialised, instances are in some state after inference completes.
Then, user clicks run button again, and same instances of classes are returned by lookups.
Oops, theese are not freshly created instances, using them as they are may cause harm.
The problem is ilustrated on fig. \ref{factory_pattern1}.
\begin{figure}
	\centering\includegraphics[scale=\myscale]{factory_pattern2}
	\caption{Factory solution of singleton lookup class problem} \label{factory_pattern2}
\end{figure}

There are basically two solutions:
\begin{itemize}
	\item force each class (submodule) that is being looked up in inference process, to implement
some sort of cleanup method, that would restart it into fresh state and make it ready to use by another 
inference run,
	\item or create new instances of theese classes in each inference run and make singleton classes (those which NetBeans return by lookups) their factory classes.
\end{itemize}
We decided for the latter approach (as ilustrated on fig. \ref{factory_pattern2}
It enables us not only to produce fresh class instance in each inference run, but also factory classes implement obligatory
module methods such as \code{getName}, \code{getDescription} and so on, which are really same in each inference run.
Each submodule is then defined by (at least) two interfaces.
One is the factory interface, like this one:
\begin{verbatim}
public interface AutomatonSimplifierFactory extends 
                     NamedModule, Capabilities, UserModuleDescription {
  <T> AutomatonSimplifier<T> create();
}
\end{verbatim}
It extends \code{NamedModule} with its modules describing methods:
\begin{verbatim}
String getName();
String getDisplayName();
String getModuleDescription();
\end{verbatim}
It extends \code{Capabilities}, thus each module have to answer, if it has some capabilities.
And, in TwoStep we often extend \code{UserModuleDescription}, which defines method:
\begin{verbatim}
String getUserModuleDescription();
\end{verbatim}
It returns description of the module, comprehensive to user, which is then displayed in properties panels.

In some circumstances, it is useful to have method \code{create()} generic.
The \jmodule{AutomatonSimplifier} works with \code{Automaton}, which itself is generic too.
But simplifying does not depend on type of symbol of automaton, so interface \code{AutomatonSimplifier<T>}
is also generic as simplifier can simplify automaton of any java type symbol.
Factory interface deals with this by defining \code{create()} method generic too.

Second interface is the one, which is returned by \code{create()} method.
That is the interface, which defines real work cycle with submodule (in example it is \code{AutomatonSimplifier<T>}).
We call this interface the worker interface of submodule.
Usage of this factory pattern follows the routine:
\begin{verbatim}
final Properties p = RunningProject.getActiveProjectProps(getName());

@@

  return ModuleSelectionHelper.lookupImpl(AutomatonSimplifierFactory.class,
            p.getProperty(PROPERTIES_AUTOMATON_SIMPLIFIER));
}
\end{verbatim}
Cluster processor \code{AutomatonMergingState} then receives factories of \code{AutomatonSimplifier} and \code{RegexpAutomatonSimplifier} submodules in its constructor.
Cluster processor then may create as many instances of submodule classes as it needs (maybe simplifying more than one automaton).
Thorough this document, we will mention only worker interface when describing submodules, since all factory interfaces are designed same way just described.

@@

// 3. process rules
final ClusterProcessor<AbstractStructuralNode> processor= clusterProcessorFactory.create();
for (Cluster<AbstractStructuralNode> cluster : clusterer.getClusters()) {
  final AbstractStructuralNode node =  processor.processCluster(clusterer, cluster.getMembers() - without sentinels);
  // 4. add to rules
  finalGrammar.add(new Element(node.getContext(),
                   node.getName(),
                   node.getMetadata(),
                   ((Element) node).getSubnodes(),
                   attList);
}
\end{verbatim}
Task of clustering is dedicated to \jmodule{Clusterer} submodule, and task of infering regular expression for each cluster is dedicated to \jmodule{ClusterProcessor} submodule. We will examine both of them.
There is also third submodule called \jmodule{RegularExpressionCleaner} actually, its purpose is just to beatify output regular expressions, no inference logic is implemented there. 
Modules currently in jInfer are drawn on fig. \ref{twostep_modules}.
\begin{figure}
	\centering\includegraphics[scale=\myscale]{twostep_modules}
	\caption{Submodules of TwoStep simplifier} \label{twostep_modules}
\end{figure}
We provide one \jmodule{Clusterer} and four \jmodule{ClusterProcessor} implementations.
Each of those will be explained further in this document.

@@

We provide \code{Clusterer} interface for classes to implement.
Its purpose is to cluster bunch of elements (rules) on input, into  bunch of cluster class instances (cluters) on output.
It has methods \code{add()} and \code{addAll()} for adding items for clustering.
Center is method \code{cluster()}, which does the clustering itself.
As it may be time-consuming operation, method throws \code{InterruptedException}.
Implementation should take care of checking whether thread is user interrupted (see \cite[p. 12]{arch}).
After clustering, implementation should hold clusters in member, as it it will be further asked by calling method \code{T getRepresentantForItem(T item)}.
Given item to this method, one can ask for representant of cluster, to which the item belongs.
If no such cluster exists (item was not added for clustering before), we recommend you throwing an exception rather than returning \code{null}.
Missing item will probably  indicate error in algorithm rather than normal workflow.
One can pull clusters from clusterer by calling \code{getClusters()} method.

@@

\subsubsection{ClustererWithAttributes extended interface}
Maybe you noticed that whole clusterer interface and cluster class are generic.
They may be used as design pattern not only for clustering elements in inference process.
To address clustering of elements in more detail, we created \code{ClustererWithAttributes<T, S>} interface, which extends \code{Clusterer<T>} interface.
It adds method \code{List<Cluster<S>{}> getAttributeClusters(T representant)}, implying that each representant of
type \code{T} (that is representant of some main cluster) has some "attribute" clusters associated with it.
Attribute clusters are of type \code{S} and can be retrieved by calling \code{getAttributeClusters(x)}.

@@

Whole submodule structure of AutomatonMergingState cluster processor is drawn on fig. \ref{automaton_merging_state_submodules}.
\begin{figure}
	\centering\includegraphics[scale=\myscale]{automaton_merging_state_submodules}
	\caption{Submodules of AutomatonMergingState cluster processor} \label{automaton_merging_state_submodules}
\end{figure}
We have implemented \emph{k,h-context} (see \cite{ahonen}) state equivalence in class \code{KHContext}, which is used by \code{Greedy}
to test mergability of states by default configuration.

\subsubsection{StateRemoval}
We are using state removal method (see \cite{1224321}) to convert regexp automaton into equivalent regular expression.
This is implemented in \code{StateRemoval} class.
We defined one submodule of this class with interface called \code{Orderer}.
It has only one method to implement: \code{getStateToRemove}.
Given automaton it has to return reference to one state which should be removed from automaton at first.
State removal calls this submodule and removes states given until there are only two states in automaton - superInitial and superFinal states,
with exactly one transition.
That transition has final regular expression on it as symbol, it is read and returned to AutomatonMergingState.
We implement one orderer, called \code{Weighted}.
It is simple heuristic - weights all states (weight = sum of {in | out | loop}-transition regular expression lengths) and returns state with lowest weight.
